---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
I read the data and after checking the summary I did not felt a need to preprocess it.
```{r, echo = TRUE}
data = read.csv('activity.csv')
summary(data)
```


## What is mean total number of steps taken per day?
The total number of steps taken per day is
```{r, echo = TRUE}
library(dplyr)
stepsPerDay = summarise(
            group_by(
              arrange(
                select(
                  filter(data, !is.na(steps)),
                  steps, date),
                date),
              date),
            totalSteps = sum(steps))
stepsPerDay
```

A barplot of our data (total number of steps per day) would show a number of steps for each day. A histogram would show a distribution of a number of steps per day as below
```{r, echo = TRUE}
library(ggplot2)
ggplot(stepsPerDay, aes(x=totalSteps)) + geom_histogram(colour = 'blue', fill = 'blue')
```

The mean of the total number of steps per day is
```{r, echo = TRUE}
mean(stepsPerDay$totalSteps)
```

And the median of the total number of steps per day is
```{r, echo = TRUE}
median(stepsPerDay$totalSteps)
```


## What is the average daily activity pattern?
The average daily activity pattern can be seen in a time series plot below. After calculating the average number of steps in a five minute interval across all days I also transform notion of five time intervals into an hour:minute format. In the end I made a plot.
```{r, echo = TRUE}
fiveMins = summarise(
            group_by(
              arrange(data, interval),
              interval),
            steps = mean(steps, na.rm = TRUE))
mns <- function(m) {
  x <- m * 60
  return(x)
}
start = strptime('00:00', '%H:%M')
fiveMins$interval = start + mns(fiveMins$interval)
library(scales)
ggplot(fiveMins, aes(interval, steps)) + geom_line() + scale_x_datetime(labels = date_format('%H:%M')) + xlab('Five minute intervals in a day') + ylab('Average number of steps')
```

The maximum number of steps on average across all the days in the dataset have the below 5-minute interval
```{r, echo = TRUE}
fiveMins = summarise(
  group_by(
    arrange(data, interval),
    interval),
  steps = mean(steps, na.rm = TRUE))
fiveMins$interval[fiveMins$steps == max(fiveMins$steps)]
```


## Imputing missing values
The total number of missing values in the dataset is
```{r, echo = TRUE}
nrow(data[is.na(data),])
```

I devised a strategy for filling in all of the missing data in the dataset: replacing them with the mean value in a five minute interval across all days. I stored the result in a new dataset.
```{r, echo = TRUE}
noNA = data
for (i in 1:nrow(noNA)) {
  if (is.na(noNA$steps[i])) {
    noNA$steps[i] = fiveMins$steps[fiveMins$interval == noNA$interval[i]]
  }
}
```

The histogram of the total number of steps taken each day is as below
```{r, echo = TRUE}
stepsPerDay1 = summarise(
  group_by(
    arrange(
      select(
        filter(noNA, !is.na(steps)),
        steps, date),
      date),
    date),
  totalSteps = sum(steps))

ggplot(stepsPerDay1, aes(x=totalSteps)) + geom_histogram(colour = 'blue', fill = 'blue')
```

The mean value is
```{r, echo = TRUE}
mean(stepsPerDay1$totalSteps)
```

And the median value is
```{r, echo = TRUE}
median(stepsPerDay1$totalSteps)
```

Only the median differ from original values. Imputing missing data on the estimates of the total daily number of steps enlarge the median value.

## Are there differences in activity patterns between weekdays and weekends?
I added a new factor variable indicating if a day is a weekday or a weekend.
```{r, echo = TRUE}
noNA$week = weekdays(as.Date(noNA$date))
noNA$week[noNA$week == 'ponedeljek'] = 'weekday'
noNA$week[noNA$week == 'torek'] = 'weekday'
noNA$week[noNA$week == 'sreda'] = 'weekday'
noNA$week[noNA$week == 'Äetrtek'] = 'weekday'
noNA$week[noNA$week == 'petek'] = 'weekday'
noNA$week[noNA$week == 'sobota'] = 'weekend'
noNA$week[noNA$week == 'nedelja'] = 'weekend'
noNA$week = factor(noNA$week)
```

Differences in activity patterns between weekdays and weekends are clearly noted in a panel plot below.
```{r, echo = TRUE}
fiveMins1 = summarise(
  group_by(
    arrange(noNA, week, interval),
    week, interval),
  steps = mean(steps))

ggplot(fiveMins1, aes(interval, steps)) + facet_wrap( ~ week, ncol = 1) + geom_line() + ylab('Average number of steps')
```